### 2025-04-27 conteudo/tool/architecture.tex
fix architecture diagram
reforce that we do not expect good results in T3/T4 types
update some things on the tool

The Setup step consists of the Function Breaker and the Code Duplication Finder. The pipeline of this step works as follows: the Function Parser receives the codebase the user is interested in, extracts the functions of the codebase along with metadata, and creates a new temporary codebase where the functions extracted become new code files. The Code Duplication Finder iterates over every pair of files in the temporary codebase, checks if they are code clones, and, if so, saves them in the Code Duplication Database, which is a text file that stores every code duplication as a triple \textbf{<function1,function2,similarity>}, where \textbf{function1} and \textbf{function2} are the functions that are duplicates of each other, and \textbf{similarity} is the metric given by the code duplication detection method utilized, which in our case is the cosine similarity explained in \ref{sec:similarity}. Finally, the Query Responder consumes the temporary codebase and the Code Duplication Database to extract duplicated functions-related information per user request. Figure \ref{fig:diagram} illustrates the tool pipeline. An explanation of how each component works is provided in this section.

\begin{figure}
\includegraphics{diagrama_mestrado}
\caption{Architecture diagram demonstrating the relationship between the tool components.}
\label{fig:diagram}
\end{figure}

@@

\begin{itemize}
	\item \textbf{C}: Given a source code file, we extract every code block from it, maintaining the depth of the code block in the code context and some of the code lines before the definition of the code block. According to the C programming language's grammar, functions are defined in a global context at depth 0. There are other elements with the same depth as functions, and we differentiate them by analyzing if the first non-empty character before the curly brackets of the code block is a closing parenthesis. This pattern uniquely identifies functions within C code.

	\item \textbf{Java}:  Given a source code file, we extract every code block from it, maintaining the depth of the code block in the code context and some of the code lines before the definition of the code block. According to Java's grammar, functions are defined inside classes. Functions can exist at different depths, as Java accepts inner classes. We differentiate functions from other elements by analyzing if the first non-empty character before the curly brackets of the code block is a closing parenthesis. A challenge with this approach in Java is that other elements, such as for-loops, can follow this pattern, necessitating additional filters. We chose to analyze functions at depth 1, which ignores any other element that is a code block and has a parenthesis as the first non-empty character before the curly brackets. This choice means the extractor does not find functions declared inside inner classes, but this limitation is not an issue in this work, as there are no inner classes in BigCloneBench \citep{bigclonebench}, the only Java codebase explored in this work.
\end{itemize}

@@

@@

For the code duplication detection method, we treat the source code files as text and apply the TF-IDF vector embedding method implemented by the Gensim library \citep{gensim}, then compute cosine similarity as the similarity metric. This method was chosen for its claimed performance, programming language independence, and the fact that it does not require compilable code, which is expected in the temporary codebase, as it does not contain complete code artifacts. \cite{platistool} implemented this code duplication detection method and released it under the MIT license \citep{mitlicense}. We adapted his implementation to fit our tool's expected input and output formats. Switching the duplication detection method is feasible, similar to how we adapted Plates' implementation.

@@

\subsection{Setup step}
\label{subsec:setup}

The Setup step is a procedure executed by the user once per codebase. The Setup step takes the input codebase and a minimum similarity metric value for function pairs considered duplicates. It then runs the Function Parser and the Code Duplication Finder to create a temporary codebase and the Code Duplication Database. Setting a minimum similarity metric reduces the Code Duplication Database size, optimizing memory usage and the computational cost of the Query Responder step. For example, if this parameter does not exist and we work with a codebase of $10000$ functions, each with a 50-character relative path and function name, the Code Duplication Database would reach approximately $5000^2 \times 50 ~= 5$ gigabytes. As most function pairs are unlikely to be duplicated in large codebases, this file size can be significantly reduced.

\subsection{Query Responder step}

@@

\begin{itemize}
	\item Number of duplicated functions above a similarity threshold.
	\item Number of duplicated functions equal or above a similarity threshold.
	\item List of duplicated functions above a similarity threshold.
	\item List of duplicated functions equal to or above a similarity threshold.
\end{itemize}

>>>>

The Preprocessor contains two main components, the Function Breaker and the Code Duplication Finder. 
The flow of how the Preprocessor works is as follow: Function Parser receives the codebase the user is interested in,
extracts the functions of the codebase along with metadata, and creates a new temporary codebase where the functions extracted become new code files. 
The Code Duplication Finder iterates over every pair of files in the Temporary Codebase, checks if they are code clones, and, 
if so, saves them in the Code Duplication Database, which is a text file that stores every code duplication as a triple 
\textbf{<function1,function2,similarity>}, where \textbf{function1} and \textbf{function2} are the functions that are duplicates 
of each other, and \textbf{similarity} is the metric given by the code duplication detection method utilized in the Code Duplication Finder. ,

The Query Responder consumes the Temporary Codebase and the Code Duplication Database to extract duplicated functions-related information per user request.

Figure \ref{fig:diagram} illustrates the tool architecture. An explanation of how each component works is provided in this section.

\begin{figure}
\includegraphics[scale=0.5]{diagrama_mestrado}
\caption{Architecture diagram demonstrating the relationship between the tool components.}
\label{fig:diagram}
\end{figure}

@@

\subsubsection{C Programming Language}

For a source file in the C programming language, we extract the code blocks with depth 0, as per the programming 
language's grammar, functions are defined in a global context at depth 0. To extract the code blocks, we firstly need to 
identify the portions of code that are comments or \textit{\#define} directives and we mark them as irrelevant, after that, 
it is simple work of finding the curly brackets that are in depth 0, and with that we find the body of the functions.

To extract the header of the functions, we start at the open curly brackets and go back in the source file consireding 
the programming language grammar  and the portions of code that have been marked as irrelevant because of comments and 
\textit{\#define} directives.

There are other elements in depth 0 that are not functions, we analyze if the code blocks is of a function at the same 
we try to extract the header of the code block.

\subsubsection{Java Programming Language}

For the Java programming language, we have decided to implemente a simpler approach that does not support 
the programming language completely, as the only Java codebase explored in this work is the BigCloneBench \citep{bigclonebench} and 
the approach implemented cover the needs of this codebase.

For a source file in Java, we extract the code blocks with depth 1 and some code lines before the definition of the code block. 
Then, we check if the code blocks are functions by analyzing if the first non-empty character before the curly brackets of 
the code block is a closing parenthesis.

As per Java's grammar, functions are defined inside classes and can exist at different depths, as Java accepts inner classes, 
that means that this approach does not work with inner classes. The challenge to implement a more generic approach 
is that other elements of depth 2 or deeper, such as for-loops, can follow the closing parenthesis pattern, necessitating 
a better analysis to different these elements. The BigCloneBench \citep{bigclonebench} does not contains inner classes.

@@

\label{subsec:code_finder}

@@

We implemented two methods for code duplication detection in the tool, in which the user can choose what to use.
The first method is based in text similarity, the second is based in simpler method based on the number of equal lines.
The experiments and tests on this research were done using the text similarity method, the purpose of the 
second method is to demonstrate that the tool can have multiple duplication detection methods.

\subsubsection{Text Similarity Method}

For the text similarity method, we treat the source code files 
as text and apply the TF-IDF vector embedding method implemented by the 
Gensim library \citep{gensim}, then compute cosine similarity as the 
similarity metric.

This method was chosen for its claimed performance, 
programming language independence, and the fact that it does not require 
compilable code, which is expected in the temporary codebase, as it does 
not contain complete code artifacts. We expect our method to yield good 
results for Type-1 and Type-2 code duplication types while performing 
poorly for the other types. It is possible to change the code duplication 
detection method to one of the state-of-the-art methods if needed.

\cite{platistool} implemented this code duplication detection method and 
released it under the MIT license \citep{mitlicense}. We adapted his 
implementation to fit our tool's expected input and output formats. 
Switching the duplication detection method is feasible, similar to how 
we adapted Plates' implementation.

\subsubsection{Difference Method}

For the difference method, 
we implemented a method considering the number of exactly equal lines between two functions.
For two functions \textit{function1} and \textit{function2}, let $NL_1$ be the number of lines in \textit{function1},
$NL_2$ be the number of lines in \textit{function2}, and $EQ$ be the number of equal lines between \textit{function1} and
\textit{function2}, we compute the similarity with the formula:

$$\frac{2 \cdot EQ}{NL_1 + NL_2}$$

This code duplication method is considerably simpler and more explainable then the text similarity method, 
as the metric is the ratio of common lines between the two functions. To compute the number of equal lines, we use
the \textit{diff command} built-in in the Linux environment \citep{diffcommand}.

The difference method was not tested on the evaluation of the tool and other phases of this research. 
The method was implemented to show that tool can have multiple duplication detection methods.

@@

\subsection{Preprocessor}
\label{subsec:setup}

The Preprocessor is a procedure executed by the user once per codebase. 
The Preprocessor takes the input codebase and a minimum similarity metric value for function pairs considered duplicates. 
It then runs the Function Parser and the Code Duplication Finder to create a temporary codebase and the Code Duplication Database. 
Setting a minimum similarity metric reduces the Code Duplication Database size, optimizing memory usage and the computational 
cost of the Query Responder step. For example, if this parameter does not exist and we work with a codebase of $10000$ functions, 
each with a 50-character relative path and function name, the Code Duplication Database would reach approximately $5000^2 \times 50 ~= 5$ gigabytes. 
As most function pairs are unlikely to be duplicated in large codebases, this file size can be significantly reduced.

\subsection{Query Responder}

\todo[inline]{I NEED TO CHANGE HERE}

@@

\begin{itemize}
	\item Number of dupliddcated functions above a similarity threshold.
	\item Number of duplicated functions equal or above a similarity threshold.
	\item List of duplicated functions above a similarity threshold.
	\item List of duplicated functions equal to or above a similarity threshold.
\end{itemize}

### 2025-04-27 conteudo/tool/tool.tex
update some things on the tool

The tool operates in two main steps. The first step, called \textbf{Setup}, is
executed only once for each codebase. This step processes the codebase by
extracting functions from code files and identifying code clone duplications.
The second step, called \textbf{Query Responder}, utilizes the information obtained in
the Setup phase to execute the tool functionalities as needed. Detailed
descriptions of the Setup and Query Responder processes can be found in
Sections \ref{subsec:architecture} and \ref{subsec:func}.

>>>>

The tool operates in two main parts, the \textbf{Preprocessor} and the \textbf{Query Responder}. 
The \textbf{Preprocessor} is responsible to perform the heavy computations to find the code duplications 
accross a codebase and produce artifact to consulte information related to the duplications in a structured form.
The \textbf{Query Responder} consumes the artifacts produced by the Preprocessor to execute the tool functionalites as requested.
This choice enable the tool to concentrate the heavy and slows parts that can be execute only once per codebase, 
thus allowing a fast performance for multiple queries related to a single codebase.
More detailed descriptions of the Setup and Query Responder processes can be found in
Sections \ref{subsec:architecture} and \ref{subsec:func}.
